---
title: "US commercial flights analysis - Summary report"
output:
  html_document:
    df_print: paged
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.show = 'asis')
knitr::opts_chunk$set(dev = 'png')
knitr::opts_chunk$set(results = 'asis')
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(RSQLite)
library(ggplot2)
library(tidyr)
library(dplyr)
library(caret)
library(lmvar)
library(GGally)
library(scales)
library(ggpubr)
library(tidyquant)
library(zoo)
setwd("/Users/davidemastricci/Desktop/Progetto_Statistica/data")
```


# Introduction

Analyzig the data from  **Research and Innovative Technology Administration (RITA)** we were able to answer the following questions:


1. **How is the trend of commercial flights in the US?**



2. **Which is the best (worst) day of the week to take a flight?**

3. **Which is the best (worst) season to flight in terms of delay?**


# Results

The insight we found are:

1. The trend of commercial flights in the USA as a number of flights seems **to grow**  year by year, even though in a non-linear way.

```{r}
path_db = "/Users/davidemastricci/Desktop/Progetto_Statistica/data/ontime.sqlite3"
ontime <- dbConnect(RSQLite::SQLite(), dbname=path_db)
from_db <- function(sql) {
  dbGetQuery(ontime, sql)
}
```

```{r warning=FALSE, message=FALSE}

no_cancelled_flights <- from_db(" select Year, Month, DayofMonth, DayOfWeek, ArrDelay
                    from ontime
                    where Cancelled <> 1 and Year > 1987 ") #1987 has only three months recorded
# Adjusting the type of some variables
no_cancelled_flights$Month <- as.factor(no_cancelled_flights$Month)
no_cancelled_flights$DayOfWeek <- as.factor(no_cancelled_flights$DayOfWeek)

no_cancelled_flights$DayOfWeek <- recode(no_cancelled_flights$DayOfWeek,
       "1"="Monday",
       "2"="Tuesday",
       "3"="Wednesday",
       "4"="Thursday",
       "5"="Friday",
       "6"="Saturday",
       "7"="Sunday")

no_cancelled_flights$Month<- recode(no_cancelled_flights$Month,
                      "1"= "Jan", 
                      "2"= "Feb",
                      "3"= "Mar", 
                      "4"= "Apr",
                      "5"= "May",
                      "6"= "Jun",
                      "7"= "Jul",
                      "8"= "Aug",
                      "9"= "Sep",
                      "10"= "Oct",
                      "11"= "Nov",
                      "12"= "Dec")

```

```{r echo=FALSE}
number_of_flights_analysis <- no_cancelled_flights %>%
  group_by(Year) %>%
  summarise(NumberOfFlights= n()) # create data frame grouped by year and the number of flight for each year

```

```{r}

loess_plot <- ggplot(number_of_flights_analysis, aes(x=Year, y=NumberOfFlights))+
  geom_line()+
  stat_smooth(method = 'loess', aes(colour = 'loess'), level = .95, span = .3) +
  ggtitle("Flight trend in US")+
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))

loess_plot
```

2. The best day of the week to fly which minimize the probability of having a dalayed flight is **Saturday** with an average of percentage of delayed flights that moves from 30% to 52%. The warse day of the week are both **Friday** and **Thursday** with an avareges of percentage of delayed flights that moves respectively between 40% and 62% (Friday) and 39% and 63% (Thursday).


```{r}
dayOfWeek_analysis <- no_cancelled_flights %>%
  group_by(Year,DayOfWeek) %>%
  summarise(TotalDelay= sum(ArrDelay)) # create data frame grouped by year and the number of flight for each year


ggplot(dayOfWeek_analysis, aes(x=Year, y=TotalDelay, color=DayOfWeek))+
  geom_line()+
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(color = "Day of Week")


```

```{r echo=FALSE}

no_cancelled_flights$Month<- recode(no_cancelled_flights$Month,
                      "Jan" = "1", 
                      "Feb" = "2",
                      "Mar" = "3", 
                      "Apr"="4",
                      "May"="5",
                      "Jun"="6",
                      "Jul"="7",
                      "Aug"="8",
                      "Sep"="9",
                      "Oct"="10",
                      "Nov"="11",
                      "Dec"="12")


no_cancelled_flights <- no_cancelled_flights %>%
  mutate(Date = paste0(DayofMonth, "-",Month,"-",Year))

no_cancelled_flights$Date <- as.Date(no_cancelled_flights$Date, "%d-%m-%Y")

late_flight <- no_cancelled_flights %>%
  filter(ArrDelay > 0)

delay_analysis <- late_flight %>%
  group_by(Date) %>%
  summarise(NumberOfFlights = n())

temp <- no_cancelled_flights %>%
  group_by(Date) %>%
  summarise(NumberOfFlights = n())

delay_analysis <- full_join(temp, delay_analysis, by="Date")
delay_analysis$PercLateFlight <- (delay_analysis$NumberOfFlights.y / delay_analysis$NumberOfFlights.x)
# Days with no delayed flight will have NA as a value of column PercLateFlight so here we subistute with 0 which
# means 0% of delayed flights
delay_analysis$PercLateFlight[is.na(delay_analysis$PercLateFlight)] <- 0

moving_average_30 <- rollmean(delay_analysis$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)

delay_analysis_day_week <- late_flight %>%
  group_by(Date, DayOfWeek) %>%
  summarise(NumberOfFlights = n())

temp <- no_cancelled_flights %>%
  group_by(Date, DayOfWeek) %>%
  summarise(NumberOfFlights = n())

delay_analysis_day_week <- full_join(temp, delay_analysis_day_week, by="Date")
delay_analysis_day_week$PercLateFlight <- (delay_analysis_day_week$NumberOfFlights.y / delay_analysis_day_week$NumberOfFlights.x)
# Days with no delayed flight will have NA as a value of column PercLateFlight so here we subistute with 0 which
# means 0% of delayed flights
delay_analysis_day_week$PercLateFlight[is.na(delay_analysis_day_week$PercLateFlight)] <- 0

monday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Monday",]
tuesday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Tuesday",]
wednesday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Wednesday",]
thursday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Thursday",]
friday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Friday",]
saturday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Saturday",]
sunday <- delay_analysis_day_week[delay_analysis_day_week$DayOfWeek.x == "Sunday",]




############# THURSDAY PLOT #####################
moving_average_30 <- rollmean(thursday$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
thursday_plot <- thursday %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("THURSDAY")

############# FRIDAY PLOT #####################
moving_average_30 <- rollmean(friday$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
friday_plot <- friday %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("FRIDAY")

############# SATURDAY PLOT #####################
moving_average_30 <- rollmean(saturday$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
saturday_plot <- saturday %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("SATURDAY")



thursday_ma <- rollmean(thursday$PercLateFlight, 30)
thursday_ma.max <- max(thursday_ma)
thursday_ma.min <- min(thursday_ma)
friday_ma <- rollmean(friday$PercLateFlight, 30)
friday_ma.max <- max(friday_ma)
friday_ma.min <- min(friday_ma)
saturday_ma <- rollmean(saturday$PercLateFlight, 30)
saturday_ma.max <- max(saturday_ma)
saturday_ma.min <- min(saturday_ma)

saturday_plot;
friday_plot;
thursday_plot;
```

3. The best season to fly to minimize the probability of having a dalayed flight is **Fall** with an average of percentage of delayed flights that goes from 29% to 57%. The worst seasons, instead, are **Summer** and **Winter** with an avareges of percentage of delayed flights that moves respectively between 36% and 63% (Summer) and 33% and 66% (Winter).


```{r}

no_cancelled_flights$Month <- as.factor(no_cancelled_flights$Month)
no_cancelled_flights$Month<- recode(no_cancelled_flights$Month,
                      "1"= "Jan", 
                      "2"= "Feb",
                      "3"= "Mar", 
                      "4"= "Apr",
                      "5"= "May",
                      "6"= "Jun",
                      "7"= "Jul",
                      "8"= "Aug",
                      "9"= "Sep",
                      "10"= "Oct",
                      "11"= "Nov",
                      "12"= "Dec")

no_cancelled_flights$Season <- no_cancelled_flights$Month
no_cancelled_flights$Season<- recode(no_cancelled_flights$Season,
                      "Jan" = "Winter", 
                      "Feb" = "Winter",
                      "Mar" = "Spring", 
                      "Apr" = "Spring",
                      "May" = "Spring",
                      "Jun" = "Summer",
                      "Jul" = "Summer",
                      "Aug" = "Summer",
                      "Sep" = "Fall",
                      "Oct" = "Fall",
                      "Nov" = "Fall",
                      "Dec" = "Winter")


season_analysis <- no_cancelled_flights %>%
  group_by(Year,Season) %>%
  summarise(TotalDelay= sum(ArrDelay)) # create data frame grouped by year and the number of flight for each year


ggplot(season_analysis, aes(x=Year, y=TotalDelay, color=Season))+
  geom_line()+
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(color = "Season")

```

```{r }
late_flight <- no_cancelled_flights %>%
  filter(ArrDelay > 0)

delay_analysis_season <- late_flight %>%
  group_by(Date, Season) %>%
  summarise(NumberOfFlights = n())

temp <- no_cancelled_flights %>%
  group_by(Date, Season) %>%
  summarise(NumberOfFlights = n())

delay_analysis_season <- full_join(temp, delay_analysis_season, by="Date")
delay_analysis_season$PercLateFlight <- (delay_analysis_season$NumberOfFlights.y / delay_analysis_season$NumberOfFlights.x)
# Days with no delayed flight will have NA as a value of column PercLateFlight so here we subistute with 0 which
# means 0% of delayed flights
delay_analysis_season$PercLateFlight[is.na(delay_analysis_season$PercLateFlight)] <- 0


winter <- delay_analysis_season[delay_analysis_season$Season.x == "Winter",]
fall <- delay_analysis_season[delay_analysis_season$Season.x == "Fall",]
summer <- delay_analysis_season[delay_analysis_season$Season.x == "Summer",]
spring <- delay_analysis_season[delay_analysis_season$Season.x == "Spring", ]


############# WINTER PLOT #####################
moving_average_30 <- rollmean(winter$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
winter_plot <- winter %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("WINTER")

############# FALL PLOT #####################
moving_average_30 <- rollmean(fall$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
fall_plot <- fall %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("FALL")


############# SUMMER PLOT #####################
moving_average_30 <- rollmean(summer$PercLateFlight, 30)
min_ma <- min(moving_average_30)
max_ma <- max(moving_average_30)
summer_plot <- summer %>%
  ggplot(aes(x=Date, y=PercLateFlight))+
  geom_line()+
  geom_hline(yintercept = min_ma, color='red')+
  geom_hline(yintercept = max_ma, color='red')+
  geom_hline(yintercept = .5, color='green')+
  geom_ma(aes(color="MA 30"),ma_fun = SMA, n=30) +
  scale_y_continuous(labels = scales::format_format(scientific= FALSE, big.mark=".", decimal.mark=","))+
  labs(y="% of delayed flights", x='Timeline')+
  ggtitle("SUMMER")






fall_plot;
summer_plot;
winter_plot
```
